---
title: "P8105 Homework 2"
author: "Jenna Mohammed"
date: '2023-10-04'
output: github_document
---

```{r setup, message=FALSE}
library(dplyr)
library(tidyr)
library(tidyverse)
library(readxl)
```

## Problem #1


```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols_df = read_csv("./fivethirtyeight_datasets/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"),sep = "-") %>%
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  select(-prez_dem, -prez_gop, -day)
    
   
snp = 
  read_csv("./fivethirtyeight_datasets/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 

unemployment = 
  read_csv("./fivethirtyeight_datasets/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) 
  
```

## Problem #2

#### Read in each sheet from the trash wheel dataset. Clean each sheet by using the janitor function and removing unnecessary rows. Use the mutate function to create a new variable to determine the amount of homes that are powered by the trash found in each trash wheel.

```{r trash wheel}

# Mr Trash Wheel

mtw_sheet = read_excel("./trashwheel_data.xlsx",
  sheet = "Mr. Trash Wheel",
  skip = 1) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
    mutate(
      homes_powered = (weight_tons * 500)/30
    )
  
# Professor Trash Wheel
ptw_sheet = read_excel("./trashwheel_data.xlsx",
  sheet = "Professor Trash Wheel", 
  skip = 1) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
    mutate(
      homes_powered = (weight_tons * 500)/30
    )
                       
print(ptw_sheet)

# Captain Trash Wheel

ctw_sheet = read_excel("./trashwheel_data.xlsx",
  sheet = "Captain Trash Wheel",
  skip = 1) |>
  janitor::clean_names()|>
  drop_na(dumpster) |>
    mutate(
      homes_powered = (weight_tons * 500)/30
    )

print(ctw_sheet)

# Gwynnda Trash Wheel

gtw_sheet = read_excel("./trashwheel_data.xlsx",
  sheet = "Gwynnda Trash Wheel", 
  skip = 1) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
    mutate(
      homes_powered = (weight_tons * 500)/30
    )

print(gtw_sheet)
```

#### Combining the three datasets

```{r combining trash wheel data}
total_tw_data = 
  bind_rows(ptw_sheet, gtw_sheet, ctw_sheet)

print(total_tw_data) 
      
```

Description: After combining the three datasets, there are 289 observations and 13 variables. Some Key variables are the weight and volume collected, homes powered, and the different types of trash collected. Professor Trash Wheel collected 216.126 tons of of trash. In July 2021, Gwynnda collected 16300 cigarette butts.

The `Mr. Trash Wheel Sheet` has `r nrow(mtw_sheet)` observations and `r ncol(mtw_sheet)` columns

The `Professor Trash Wheel` Sheet has `r nrow(ptw_sheet)` observations and `r ncol(ptw_sheet)` columns

The `Gwynnda Trash Wheel` Sheet has `r nrow(gtw_sheet)` observations and `r ncol(gtw_sheet)` columns

The Combined trash wheel data has `r nrow(total_tw_data)` observations and `r ncol(total_tw_data)` columns


## Problem 3 

#### Loading in MCI Baseline data and cleaning by removing ineligible participants. Using the mutate function to change the numerical values to categorical ones.

```{r mci baseline}

# Changing sex and APOE4 carrier status from numerical to categorical

mci_baseline_df = read_csv("./data_mci./data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names()|>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        0 ~ "female"),
    sex = as.factor(sex)) |>
  mutate(
    apoe4 = 
      case_match(
        apoe4, 
        1 ~ "carrier", 
        0 ~ "non-carier"),
    apoe4 = as.factor(apoe4)) |>
    
# remove any participants who do not meet the stated inclusion criteria 

filter(age_at_onset == "."| age_at_onset > current_age)

```

#### Finding the mean baseline age

```{r mci baseline mean}
mean(mci_baseline_df$current_age)
```

#### Determining the proportion of women in the study that are APOE4 carriers

```{r mci baseline female carrier}
data_female = filter(mci_baseline_df, sex == "female")
nrow(data_female)

data_female_carrier = filter(data_female, apoe4 == "carrier")
nrow(data_female_carrier)
```

#### Determining how many participants developed MCI

```{r}

filter(mci_baseline_df, age_at_onset !=".") |> nrow()
```


MCI Baseline description: Important steps of this section were removing the first row of the data to ensure the variables would be read properly. Another is changing the sex and apoe4 variables from numeric to categorical using the mutate function.

There are 479 participants in this dataset, and from that, 93 of them developed MCI. The average baseline age is 65.03 . The proportion of women in the study that are APOE4 carriers is 63/479.



#### Loading in and cleaning MCI Longtitudinal data by using the janitor function.

```{r, longitudinally observed data}

mci_amyloid_df = read_csv("./data_mci./data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names()|>
  mutate()
```

#### Combining demographic and biomarker datasets

```{r}

colnames(mci_amyloid_df)[colnames(mci_amyloid_df) == "study_id"] = "id"

mci_combined_df = full_join(mci_baseline_df, mci_amyloid_df, by = c("id"))

view(mci_combined_df)
```

MCI Longtitudal Description: An important apsect of this section was removing the first row of the data to ensure the variables would be read properly. Secondly, when joining the amyloid data and the longtitudal data, we needed to change the column `study_id` to `id`. 

#### Exporting the combined dataset as a csv file

```{r save csv}
write.csv(mci_combined_df, "./Combined_MCI.csv")
```

